What is an Ancestral Recombination Graph?

### 2022-02-17 paper.tex
Another pass through the Event ARG section

The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination on a single
chromosome, but it has significant drawbacks as the basis for a general purpose
storage and interchange format. To avoid confusion, we will refer to this
ARG encoding as an ``Event ARG'', or EARG.
There are many ways in which the EARG structure can be formalised (see
e.g.~\citep{gusfield2014recombinatorics}).
% It would be slightly easier for us if we defined nodes as in N_0, since
% we use 0 as a node ID, but leaving for now.
We can define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
list of tuples such that $e_j = (c, p)$ defines a directed edge between
child node $c$ and parent $p$ ($c, p \in \mathbb{N}$), and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}_0$ is a function
mapping nodes to recombination breakpoints, such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
% This is arbitrary, but handy when you start defining the function
% for recovering trees; see below
$\sigma(u) = \infty$ otherwise. For each recombination node $u$ we must
have exactly two edges in $E$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.
An example is given in Fig.~\ref{fig-arg-data-structure}A.

@@

% Not sure if we actually include this, but it's good to work it through
% to make sure the notation works.
Given a set of sample nodes $S$ we can recover the local tree at
position $x$ by the following procedure. Let $N$ be a set of
nodes we are currently constructing the tree at, and let
$p: \mathbb{N} \rightarrow \mathbb{N}$ be the mapping from child
node to parent. Firstly, initialise $N \leftarrow S$.
Then, while $|N| > 0$, we add branches to the
tree as follows. Let $u$ be an arbitrary element of $N$ (which
we remove). Then, if $\sigma(u) < x$ we let $j$ be the
smallest value such that $e_{j,1} = u$, and otherwise
let $j$ be the largest value such that $e_{j,1} = u$.
Then, set $v \leftarrow e_{j, 2}$,
$p_u \leftarrow v$
and $N \leftarrow N \cup \{v\}$.

The Griffiths ARG encoding is ambiguous because it does not explicitly
define which of the parent lineages resulting from a recombination
event carries ancestral material to the left or right of the
breakpoint. The ambiguity is resolved by adopting a simple rule when
traversing the graph in order to generate the marginal trees for a
given position $z$: at a recombination node $u$, we follow the
first edge in which $u$ is a child if $f(u) \leq z$, and the
second edge otherwise. This ordering requirement, which straightforward
% Saying "awkward" too many times here, rephrase.
to describe can be awkward in practice. For example,
simulating an ARG in this representation is considerably
complicated by the fact that the
first event to occur may hit the right-hand lineage rather
than the left.
Such awkwardnesses can be worked around, of course,
but in practice methods such has
\texttt{beagle}~\citep{lyngso2005minimum} and
\texttt{KwARG}~\citep{ignatieva2021kwarg}
resolve this problem by
associating a label ``P'' (prefix) or ``S'' (suffix) with the
outbound edges of a recombination event.

>>>>

The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination on a single
chromosome, but it has significant drawbacks as the basis for a general purpose
storage and interchange format. To avoid confusion, we will refer to this
ARG encoding as an ``Event ARG'', or EARG.
There are many ways in which the EARG structure can be formalised (see
e.g.~\citep{gusfield2014recombinatorics}); the following is useful
to illustrate the computational properties of the structure.
% It would be slightly easier for us if we defined nodes as in N_0, since
% we use 0 as a node ID, but leaving for now.
We define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
list of tuples such that $e_j = (c, p)$ defines a directed edge between
child node $c$ and parent $p$ ($c, p \in \mathbb{N}$), and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints, such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \infty$ otherwise. (The use of $\infty$ to denote
non-recombination nodes is arbitrary, but useful when we extract
% NOTE: need to be careful about using "local" trees if we're also
% distinguishing resolved and non-resolved GARGs as being "locally"
% defined or not. One is topologically local and the other is local
% along the genome.
local trees; see below.)
For each recombination node $u$ we must
have exactly two edges in $E$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.
An example is given in Fig.~\ref{fig-arg-data-structure}A.

@@

% % Not sure if we actually include this, but it's good to work it through
% % to make sure the notation works. Maybe something to stick into
% % an appendix
% The most fundamental operation that we need to perform on an ARG
% data structure is extracting the local tree topology at a particular
% position along the genome. The basic strategy is to traverse
% the graph upwards, building the tree as we visit each node.
% At a recombination node, when we have a choice of two parents,
% we follow the path that corresponds to the position that we
% are building the tree for. More precisely, let $S$ be the set
% of sample nodes (usually the leaves of the graph)
% and $0 \leq x < L$ be a position along the genome, and suppose
% we are building a mapping $p_u$ that defines the parent of node
% $u$ at $x$.
% Let $N$ be a set of
% nodes we are currently constructing the tree at,
% and initialise $N \leftarrow S$.
% Then, while $|N| > 0$, we add branches to the
% tree as follows. Let $u$ be an arbitrary element of $N$ (which
% we remove).
% % This isn't right: we're not handling the case where there's
% % no edges in E where the child is $u$.
% Then, if $\sigma(u) < x$ we let $j$ be the
% smallest value such that $e_{j,1} = u$, and otherwise
% let $j$ be the largest value such that $e_{j,1} = u$.
% Then, set $v \leftarrow e_{j, 2}$,
% $p_u \leftarrow v$
% and $N \leftarrow N \cup \{v\}$.

The ordering of edges is a vital element of the EARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with \emph{nodes}
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). When we want to recover
the local tree at a position $x$ along the genome,
we traverse the graph upwards from the leaves. At a particular
node $u$, if it has one parent we are at a common ancestor
node and we simply follow that parent. If we are at a
recombination node, $u$ has two parents; if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the first edge, and otherwise follow the second edge.
This ordering requirement, which although straightforward
to describe, has some drawbacks in practice. For example,
simulating an ARG in this representation is
complicated by the fact that the
first event to occur may hit the right-hand lineage rather
than the left (and so we cannot emit edges as they are generated).
Such issues can be worked around, of course,
but depending on the ordering of otherwise indistinguishable
objects is generally problematic. In practice, several
methods
explicitly associate information with the outbound edges
of a recombination event
to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.
